Comparison of MCMC algorithms in Stochastic Volatility Models

### 2023-10-05 manuscript/main.tex
updated plots and presentation guideline

    \subsection{Evaluation Metrics}
        The key metrics and diagnostics to compare the performance of these methods are the effective sample size (ESS), rank statistics, and chi-squared test statistics.

        \subsubsection{Effective sample size}
            ESS measures the efficiency of the MCMC sampler. It calculates the number of effectively independent draws from the posterior draws generated by a Markov chain. A poor ESS can arise from high autocorrelation in the Markov chain which leads to highly dependent samples. An efficient MCMC algorithm takes less resources (for example, time and number of draws) to get a representative sample of the target distribution. If a MCMC algorithm possesses higher ESS for the majority of its parameters (relative to another strategy), then there is evidence that this method is a more efficient sampler.

@@

            The shape of the rank statistic distribution gives insight into how a MCMC may be miscalibrated. Specifically, it gives information about the type of bias in the posterior estimates for a given parameter. Figures \ref{fig:overestimate}, \ref{fig:underestimation} and \ref{fig:underdispersed} gives some examples of the posterior estimates which result in bias of the rank statistics.

            Figure \ref{fig:overestimate} shows a distribution of rank statistics with a large left peak. This is consistent with posterior samples overestimating the true parameter. The sum of indicator random variables is small since the majority of posterior draws are larger than the true value. This results in a small rank statistic. Hence, the shape of the histogram is consistent with multiple posterior estimates as part of the SBC procedure being biased towards overestimation of the true parameter.

            Figure \ref{fig:underestimation} displays the converse behaviour. The large right peak is consistent with a bias towards under estimation of the true parameter.

            Figure \ref{fig:underdispersed} has large peaks on both ends of the histogram. The sampler is disproportionately over and under estimating the true parameter resulting in biased rank statistics. \citet{talts2018validating} describe this bias as underdispersion of the posterior relative to the prior distribuion (since the posterior draws should follow the same distribution as the prior in this simulation design). The estimated posterior is too narrow relative to the spread of the prior resulting in bias on both ends of the rank statistics distribution.

            \begin{figure}[H]
                \centering
                \includegraphics[scale=0.07]{methodology/lhs.png}
                \caption{Left: Non uniform rank statistics with peak on the left side. Right: Posterior distribution vertical line representing true parameter. Bias in rank statistics distribution comes from disproportionate number of SBC iterations overestimating the true parameter.}
                \label{fig:overestimate}
            \end{figure}

@@

            \begin{figure}[H]
                \centering
                \includegraphics[scale=0.07]{methodology/underdispersed.png}
                \caption{Left: Non uniform rank statistics with peaks on both ends. Right: Overlayed posterior and prior distributions for the same arbitrary parameter (taken from \citet{talts2018validating}). Bias in ranks comes from underdispersed posterior distribution relative to prior.}
                \label{fig:underdispersed}
            \end{figure}

@@

>>>>

    \subsection{Evaluation Metrics}
        The key metrics and diagnostics to compare the performance of these methods are rank statistics, and chi-squared test statistics to measure calibration and the effective sample size (ESS) to measure effciency.

@@

            The shape of the rank statistic distribution gives insight into how a MCMC may be miscalibrated. Specifically, it gives information about the type of bias in the posterior estimates for a given parameter. Figures \ref{fig:underestimation} and \ref{fig:underdispersed} are recreated from \citet{talts2018validating} for conveience which gives some examples of the posterior estimates which result in bias of the rank statistics.

            Figure \ref{fig:underestimation} shows a distribution of rank statistics with a large right peak. This is consistent with posterior samples underestimating the true parameter, where the estimated posterior distrubtion is biased to the left of the prior. The sum of indicator random variables is large since the majority of posterior draws are smaller than the true value. This results in a large rank statistic and a bias on the right side of the histogram. The coverse is true if the peak of the histogram is on the left hand side (i.e the posterior is biased towards the right hand side of the prior).

            Figure \ref{fig:underdispersed} has large peaks on both ends of the histogram. The sampler is disproportionately over and under estimating the true parameter. \citet{talts2018validating} describe this bias as underdispersion of the posterior relative to the prior distribuion. The estimated posterior is too narrow relative to the spread of the prior resulting in bias on both ends of the rank statistics distribution.

            % \begin{figure}[H]
            %     \centering
            %     \includegraphics[scale=0.07]{methodology/lhs.png}
            %     \caption{Left: Non uniform rank statistics with peak on the left side. Right: Posterior distribution vertical line representing true parameter. Bias in rank statistics distribution comes from disproportionate number of SBC iterations overestimating the true parameter.}
            %     \label{fig:overestimate}
            % \end{figure}

@@

            \begin{figure}[H]
                \centering
                \includegraphics[scale=0.07]{methodology/underdispersed.png}
                \caption{Left: Non uniform rank statistics with peaks on both ends. Right: Overlayed posterior and prior distributions for the same arbitrary parameter. Bias in ranks comes from underdispersed posterior distribution relative to prior.}
                \label{fig:underdispersed}
            \end{figure}

@@

            \subsubsection{Effective sample size}
            ESS measures the efficiency of the MCMC sampler. It calculates the number of effectively independent draws generated by a MCMC to estimate a target parameter. A poor ESS can arise from high autocorrelation in the Markov chain which leads to highly dependent samples. An efficient MCMC algorithm takes less resources (for example, time and number of draws) to get a representative sample of the target distribution. If a MCMC algorithm possesses higher ESS for the majority of its parameters (relative to another strategy), then there is evidence that this method is a more efficient sampler.

            The ESS is defined as $n_{eff}$ and is a function of the number of chains $M$, the number of samples per chain $N$ and the autocorrelation at lag t $\rho_t$. $T$ is a trunction lag chosen such that the sum of any two consecutive autocorrelation lags is negative. This truncation (or some kind of downweighting) is required because a large number of lags leads to noiser autocorrelation estimates. More details about this and the calculation of the autocorrelation can be found in \citep{vehtari2021rank} and \citet{geyer1992practical}.

            $$
            \begin{aligned}
                n_{eff} = \frac{NM}{1+2 \sum_{t=1}^T \hat{\rho}_t}
            \end{aligned}
            $$
